Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-10 POPL/paper.tex
build errors

\pa{We should probably define some of the words in this definition}
\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

\subsection{Parse Canonicity}
Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justify that the type
theory being studied is well-behaved. In our case, however, since we
want to connect \theoryabbv to parsers, we must provide a more
detailed account of canonicity. In particular, we give a nonstandard semantics
of \typeabbv that carries a proof of canonicity along with it.

@@

\begin{theorem}[Canonicity]
  Let $A$ be a closed, purely positive linear type. Then there is an
  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
\end{theorem}
\begin{proof}
  We outline the proof here, more details are in the appendix. The
  proof proceeds first by a standard logical families construction
  that combines canonicity arguments for dependent type theory
  \cite{coquand,etc} with logical relations constructions for linear
  types \cite{hylandschalk}. It is easy to see by induction that the
  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
  \rrbracket$ and the fundamental lemma proves that the projection
  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
  procedure. Then we establish again by induction that
  canonicalization is also a retraction by observing that introduction
  forms are taken to constructors.
\end{proof}

@@

Dual to constructing traces, we often want to construct
other terms in a context containing values of type
$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
of the elimination principle for multiple-fixed points ---
which we write as $\mathsf{mfold}$ --- given
in \cref{subsubsec:other}.

@@

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable decision procedure for DFAs in \thoeryabbv
embedded in Agda.

>>>>

\pedro{We should probably define some of the words in this definition}
\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

\subsection{Parse Canonicity}
Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justify that the type
theory being studied is well-behaved. In our case, however, since we
want to connect \theoryabbv to parsers, we must provide a more
detailed account of canonicity. In particular, we give a nonstandard semantics
of \theoryabbv that carries a proof of canonicity along with it.

@@

\begin{theorem}[Canonicity]
  Let $A$ be a closed, purely positive linear type. Then there is an
  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
\end{theorem}
\begin{proof}
  We outline the proof here, more details are in the appendix. The
  proof proceeds first by a standard logical families construction
  that combines canonicity arguments for dependent type theory
  TODO cite coquand
  % \cite{coquand,etc}
  with logical relations constructions for linear
  types
  TODO cite hylandschalk
  % \cite{hylandschalk}
  . It is easy to see by induction that the
  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
  \rrbracket$ and the fundamental lemma proves that the projection
  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
  procedure. Then we establish again by induction that
  canonicalization is also a retraction by observing that introduction
  forms are taken to constructors.
\end{proof}

@@

Dual to constructing traces, we often want to construct
other terms in a context containing values of type
$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
of the elimination principle for multiple-fixed points ---
which we write as $\mathsf{mfold}$ --- given
in \cref{fig:multifix}.

@@

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable decision procedure for DFAs in \theoryabbv
embedded in Agda.
